---
title: "Flexible Imputation of Missing Data"
author: "Stef van Buuren"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
bibliography: refs.bibtex
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE} 
knitr::opts_chunk$set(echo = TRUE, fig.height = 4, fig.width = 6, tidy = TRUE)
options(digits = 3)
set.seed(92622)
```

This page contains the R code from the book *Flexible Imputation of Missing Data* [@vanbuuren2012].

### Chapter 1 	Introduction

### Section 1.1 The problem of missing data

```{r}
### calculate the mean of three numbers
y <- c(1, 2, 4) 
mean(y)

### calculate the mean of three numbers, where one is missing
y <- c(1, 2, NA)
mean(y)

### repeat, but with any missing data removed
mean(y, na.rm=TRUE)

### store your current options (safety measure)
youroptions <- options()

### demo regression that generates an error message using default na.fail
options(na.action = na.fail)

### uncomment to see the error
### lm(Ozone ~ Wind, data=airquality)

### remove incomplete observations by 'na.action = na.omit'
fit <- lm(Ozone ~ Wind, data = airquality, na.action = na.omit)
coef(fit)

### set automatic na.action = na.omit
options(na.action = na.omit)

### find out how many rows were deleted
deleted <- na.action(fit)
naprint(deleted)

### more incomplete rows if we add Solar.R as predictor
fit2 <- lm(Ozone ~ Wind+Solar.R, data=airquality)
naprint(na.action(fit2))

### restore your original options, but set na.omit
options(youroptions)
options(na.action=na.omit)

```

### Section 1.3.2 Pairwise deletion

```{r}
### note: mean(<data.frame>) is deprecated
colMeans(airquality, na.rm=TRUE)
cor(airquality, use="pair")
cov(airquality, use="pair")
```

### Section 1.3.3 Mean imputation

```{r}
library("mice")

### impute the mean
imp <- mice(airquality, method="mean", m=1, maxit=1)

### Figure 1.1
lwd <- 1.5
par(mfrow=c(1,2))
breaks <- seq(-20, 200, 10)
nudge <- 1
x <- matrix(c(breaks-nudge, breaks+nudge), ncol=2)
obs <- airquality[,"Ozone"]
mis  <- imp$imp$Ozone[,1]
fobs <- c(hist(obs, breaks, plot=FALSE)$counts, 0)
fmis <- c(hist(mis, breaks, plot=FALSE)$counts, 0)
y <- matrix(c(fobs, fmis), ncol=2)
matplot(x, y, type="s",
        col=c(mdc(4),mdc(5)), lwd=2, lty=1,
        xlim = c(0, 170), ylim = c(0,40), yaxs = "i",
        xlab="Ozone (ppb)",
        ylab="Frequency")
box()
tp <- xyplot(imp, Ozone~Solar.R, na.groups=ici(imp),
       ylab="Ozone (ppb)", xlab="Solar Radiation (lang)",
       cex = 0.75, lex=lwd,
       ylim = c(-20, 180), xlim = c(0,350))
print(tp, newpage = FALSE, position = c(0.48,0.08,1,0.92))
```

### Section 1.3.4 Regression imputation

```{r}
fit <- lm(Ozone ~ Solar.R, data=airquality)
pred <- predict(fit, newdata=ic(airquality))

### alternative using mice
imp <- mice(airquality[,1:2], method="norm.predict", m=1, maxit=3,seed=1)

### Figure 1.2
par(mfrow = c(1,2))
fmis <- c(hist(pred, breaks, plot=FALSE)$counts, 0)
y <- matrix(c(fobs, fmis), ncol=2)
matplot(x, y, type="s",
        col=c(mdc(4),mdc(5)), lwd=2, lty=1,
        xlim = c(0, 170), ylim = c(0,40), yaxs = "i",
        xlab="Ozone (ppb)",
        ylab="Frequency")
box()
tp <- xyplot(imp, Ozone~Solar.R,
       ylab="Ozone (ppb)", xlab="Solar Radiation (lang)",
       cex = 0.75, lex=lwd,
       ylim = c(-20, 180), xlim = c(0,350))
print(tp, newpage = FALSE, position = c(0.48,0.08,1,0.92))
```

### Section 1.3.5 Stochastic regression imputation

```{r}
imp <- mice(airquality[,1:2],method="norm.nob",m=1,maxit=1,seed=1)

### Figure 1.3
par(mfrow = c(1,2))
mis  <- imp$imp$Ozone[,1]
fmis <- c(hist(mis, breaks, plot=FALSE)$counts, 0)
y <- matrix(c(fobs, fmis), ncol=2)
matplot(x, y, type="s",
        col=c(mdc(4),mdc(5)), lwd=2, lty=1,
        xlim = c(0, 170), ylim = c(0,40), yaxs = "i",
        xlab="Ozone (ppb)",
        ylab="Frequency")
box()
tp <- xyplot(imp, Ozone~Solar.R, na.groups=ici(imp),
       ylab="Ozone (ppb)", xlab="Solar Radiation (lang)",
       cex = 0.75, lex=lwd,
       ylim = c(-20, 180), xlim = c(0,350))
print(tp, newpage = FALSE, position = c(0.48,0.08,1,0.92))
```

### Section 1.3.6 LOCF and BOCF

```{r}
par(mfrow=c(1,1))
Oz <- airquality$Ozone
locf <- function(x) {
  a <- x[1]
  for (i in 2:length(x)) {
    if (is.na(x[i])) x[i] <- a
    else a <- x[i]
  }
  return(x)
}
Ozi <- locf(Oz)
colvec <- ifelse(is.na(Oz),mdc(2),mdc(1))

### Figure 1.4

plot(Ozi[1:80],col=colvec,type="l",xlab="Day number",ylab="Ozone (ppb)")
points(Ozi[1:80],col=colvec,pch=20,cex=1)
```

### Section 1.4.3 Example of multiple imputation

```{r}
imp <- mice(airquality, seed=1, print=FALSE)
fit <- with(imp, lm(Ozone ~ Wind + Temp + Solar.R))
tab <- round(summary(pool(fit)),3)
tab[,c(1:3,5)]

fit <- lm(Ozone~Wind+Temp+Solar.R,data=airquality,na.action=na.omit)
round(coef(summary(fit)),3)

### Figure 1.6
par(mfrow = c(1,2))
mis  <- imp$imp$Ozone[,1]
fmis <- c(hist(mis, breaks, plot=FALSE)$counts, 0)
y <- matrix(c(fobs, fmis), ncol=2)
matplot(x, y, type="s",
        col=c(mdc(4),mdc(5)), lwd=2, lty=1,
        xlim = c(0, 170), ylim = c(0,40), yaxs = "i",
        xlab="Ozone (ppb)",
        ylab="Frequency")
box()

tp <- xyplot(imp, Ozone~Solar.R, subset = .imp==1,
       ylab="Ozone (ppb)", xlab="Solar Radiation (lang)",
       cex = 0.75, lex=lwd,
       ylim = c(-20, 180), xlim = c(0,350))
print(tp, newpage = FALSE, position = c(0.48,0.08,1,0.92))
```

```{r}
### scatterplot of all imputed data sets (not in the book)
xyplot(imp, Ozone~Solar.R | .imp,
       ylab="Ozone (ppb)", xlab="Solar Radiation (lang)",
       cex = 0.75, lex=lwd,
       ylim = c(-20, 180), xlim = c(0,350))

### Figure 1.7

Oz <- airquality$Ozone
colvec <- ifelse(is.na(Oz),mdc(2),mdc(1))
par(mfrow=c(1,1))
plot(Oz[1:80],col=mdc(1),type="l",xlab="Day number",ylab="Ozone (ppb)")
points(Oz[1:80],col=mdc(1),pch=20,cex=1)

idx <- ici(airquality$Ozone) & (1:153)<81 
x   <- (1:153)[idx]
points(x=x,y=complete(imp,1)$Ozone[idx],col=mdc(2),pch=20,cex=1)
points(x=x,y=complete(imp,2)$Ozone[idx],col=mdc(2),pch=20,cex=1)
points(x=x,y=complete(imp,3)$Ozone[idx],col=mdc(2),pch=20,cex=1)
points(x=x,y=complete(imp,4)$Ozone[idx],col=mdc(2),pch=20,cex=1)
points(x=x,y=complete(imp,5)$Ozone[idx],col=mdc(2),pch=20,cex=1)
```

```{r}
### figure of the autocorrelation function
### (not in the book)

par(mfrow=c(2,5))
acf.ozone <- with(imp,acf(Ozone))
model <- expression(acf(resid(lm(Ozone ~ Wind + Temp + Solar.R))))
acf.resid <- with(imp,model)
calcacf <- function(acf.list) {
  k <- length(acf.list)
  acc <- acf.list[[1]]$acf
  for (i in 2:k) acc <- acc + acf.list[[i]]$acf
  return(acc/k)
 }
oz <- round(calcacf(acf.ozone$analyses),2)
re <- round(calcacf(acf.resid$analyses),2)
```


## Chapter 2 Multiple imputation

```{r}
library("mice")
library("lattice")
library("MASS")

### Section 2.1.3 The expanding literature on multiple imputation
### Figure 2.1

cit  <- c(     2010, 37, 182, NA,
               2009, 36, 129, NA,
               2008, 27, 111, NA,
               2007, 35, 136, NA,
               2006, 18, 80, NA,
               2005, 20, 76, NA,
               2004,  6, 56, NA,
               2003, 17, 42, NA,
               2002, 14, 40, NA,
               2001, 13, 36, 57,
               2000,  8, 21, 33,
               1999,  6, 25, 47,
               1998, 6, 13, 22,
               1997, 6, 18, 29,
               1996, 4, 12, 28,
               1995, 3, 5, 20,
               1994, 2, 5, 34,
               1993, 2, 6, 15,
               1991, 2, 4, 19,
               1990, 1, 2, 15,
               1989, NA, NA, 11,
               1988, 1, NA, 13,
               1987, NA, NA, 10,
               1986, NA, NA, 5,
               1985, NA, NA, 1,
               1984, NA, NA, 2,
               1983, NA, NA, 5,
               1982, NA, NA, 2,
               1981, NA, NA, 1,
               1980, NA, NA, 5,
               1979, NA, NA, 2,
               1978, NA, NA, 1,
               1977, NA, NA, 2)
cit <- matrix(cit, nr=2010-1977, nc=4, byrow=TRUE)
cit <- as.data.frame(cit)
names(cit) <- c("Year","Title","Abstract","All")
par(mfrow=c(1,1))
par(cex=0.7, lwd=0.5)
plot(x=cit$Year, y=cit$Abstract, type="o",log="y",
     xlim=c(1975,2010),ylim=c(1,200),
     ylab="Number of publications (log)", xlab="Year", pch=2, 
     axes=FALSE)
axis(1, lwd=par("lwd"))
axis(2, lwd=par("lwd"), las=1)
# box(lwd=0.5)
lines(x=cit$Year, y=cit$Title, pch=15, type="o")
lines(x=cit$Year, y=cit$All, pch=16, type="o")
legend(x=1975,y=200,legend=c("early publications",
                            "'multiple imputation' in abstract",
                            "'multiple imputation' in title"),
       pch=c(16,2,15), bty="n")
```

### Section 2.2.4 MCAR, MAR and MNAR again

```{r}
logistic <- function(x) exp(x)/(1+exp(x))
set.seed(80122)
n <- 300
y <- mvrnorm(n=n,mu=c(0,0),Sigma=matrix(c(1,0.5,0.5,1),nrow=2))
y1 <- y[,1]
y2 <- y[,2]
r2.mcar <- 1-rbinom(n, 1, 0.5)
r2.mar  <- 1-rbinom(n, 1, logistic(y1))
r2.mnar <- 1-rbinom(n, 1, logistic(y2))

### Figure 2.2

y3 <- rbind(y,y,y)
r2 <- c(r2.mcar,r2.mar,r2.mnar)
r2 <- factor(r2, labels=c("Ymis","Yobs"))
typ <- factor(rep(3:1,each=n),labels=c("MNAR","MAR","MCAR"))
d <- data.frame(y1=y3[,1],y2=y3[,2],r2=r2,typ=typ)
trellis.par.set(box.rectangle=list(col=c(mdc(2),mdc(1)),lwd=1.2))
trellis.par.set(box.umbrella=list(col=c(mdc(2),mdc(1)),lwd=1.2))
trellis.par.set(plot.symbol=list(col=mdc(3),lwd=1))
tp <- bwplot(r2~y2|typ, data=d,
             horizontal=TRUE, layout=c(1,3),
             xlab=expression(Y^2),
             col=c(mdc(2),mdc(1)),strip=FALSE, xlim=c(-3,3),
             strip.left = strip.custom(bg="grey95"))
print(tp)
```

### Section 2.3.7 Numerical example

```{r}
imp <- mice(nhanes, print=FALSE, m=10, seed=24415)
fit <- with(imp, lm(bmi~age))
est <- pool(fit)
names(est)
attach(est)
(r+2/(df+3))/(r+1)
(df+1)/(df+3)*lambda+2/(df+3)
detach(est)
```

## Chapter 3 Univariate imputation

### Section 3.1 How to generate multiple imputations

```{r fig.height = 8}
library("mice")
library("lattice")
library("gamlss")
library("rpart")

### In order to draw figure 3.1, we define two tweaked
### mice.impute.xxx functions that will store the coefficients
### of the imputation models in the global environment.
### Normally, we would not be so interested in the
### parameter estimates of the imputation model, but here
### we need them for Figure 3.1.
mice.impute.normdump <- function (y, ry, x, ...) 
{
    x <- cbind(1, as.matrix(x))
    parm <- .norm.draw(y, ry, x, ...)
    betadump <<- c(betadump,parm$beta) 
    return(x[!ry, ] %*% parm$beta + rnorm(sum(!ry)) * parm$sigma)
}
mice.impute.pmmdump <- function (y, ry, x, ...) 
{
    x <- cbind(1, as.matrix(x))
    parm <- .norm.draw(y, ry, x, ...)
    yhatobs <- x[ry, ] %*% parm$coef
    yhatmis <- x[!ry, ] %*% parm$beta
    betadump <<- c(betadump,parm$beta)
    return(apply(as.array(yhatmis), 1, .pmm.match, yhat = yhatobs, 
        y = y[ry], ...))
}


### Figure 3.1
par(mfrow=c(3,2))
data <- whiteside
lwd <- 1.5
plot(x=data$Temp, y=data$Gas, col=mdc(1), lwd=lwd, 
     xlab=expression(paste("Temperature (", degree, "C)")), 
     ylab="Gas consumption (cubic feet)")
points(x=5, y=3.6, pch=4, cex=2, lwd=lwd, col=mdc(2))
legend(x="bottomleft", legend="deleted observation", pch=4, col=mdc(2), 
       pt.lwd=lwd, bty="n", pt.cex=2)
text(x=9, y=6.5, label="a",cex=2)

data[47,"Gas"] <- NA
plot(x=data$Temp, y=data$Gas, col=mdc(1), lwd=lwd, 
     xlab=expression(paste("Temperature (", degree, "C)")), 
     ylab="Gas consumption (cubic feet)")
abline(m1<-lm(Gas~Temp, data=data, na.action=na.omit), col=mdc(4))
points(5,4.04, lwd=lwd, col=mdc(2),pch=19)
text(x=9, y=6.5, label="b",cex=2)

plot(x=data$Temp, y=data$Gas, col=mdc(1), lwd=lwd, 
     xlab=expression(paste("Temperature (", degree, "C)")), 
     ylab="Gas consumption (cubic feet)")
imp <- mice(data, m=1, maxit=0)
pred <- imp$pred
pred["Gas","Insul"] <- 0
imp <- mice(data, m=5, pred=pred, meth="norm.nob", maxit=1, print=FALSE, seed=45433)
abline(m1<-lm(Gas~Temp, data=data, na.action=na.omit), col=mdc(4))
points(rep(5,5),imp$imp$Gas, lwd=lwd, col=mdc(2),pch=19)
text(x=9, y=6.5, label="c",cex=2)

plot(x=data$Temp, y=data$Gas, col=mdc(1), lwd=lwd, 
     xlab=expression(paste("Temperature (", degree, "C)")), 
     ylab="Gas consumption (cubic feet)")
imp <- mice(data, m=1, maxit=0)
pred <- imp$pred
pred["Gas","Insul"] <- 0
betadump <- vector("list", 0) 
imp <- mice(data, m=5, pred=pred, meth="normdump", maxit=1, print=FALSE, seed=83126)
abline(m1<-lm(Gas~Temp, data=data, na.action=na.omit), col=mdc(4))
betadump <- matrix(betadump, nc=2, byrow=TRUE)
for (i in 1:5) abline(coef=unlist(betadump[i,]), col=mdc(5))
points(rep(5,5),imp$imp$Gas, lwd=lwd, col=mdc(2),pch=19)
text(x=9, y=6.5, label="d",cex=2)

pch <- c(rep(3,26),rep(1,30))
plot(x=data$Temp, y=data$Gas, col=mdc(1), lwd=lwd, pch=pch, 
     xlab=expression(paste("Temperature (", degree, "C)")), 
     ylab="Gas consumption (cubic feet)")
imp <- mice(data, m=5, meth="norm", maxit=1, print=FALSE, seed=11727)
abline(m1<-lm(Gas~Temp, data=data, na.action=na.omit, subset=Insul=="Before"), col=mdc(4))
abline(m2<-lm(Gas~Temp, data=data, na.action=na.omit, subset=Insul=="After"), col=mdc(4))
points(rep(5,5),imp$imp$Gas, lwd=lwd, col=mdc(2),pch=19)
legend(x="bottomleft", legend=c("before insulation","after insulation"), pch=c(3,1),bty="n", pt.lwd=lwd)
text(x=9, y=6.5, label="e",cex=2)

pch <- c(rep(3,26),rep(1,30))
plot(x=data$Temp, y=data$Gas, col=mdc(1), lwd=lwd, pch=pch, 
     xlab=expression(paste("Temperature (", degree, "C)")), 
     ylab="Gas consumption (cubic feet)")
betadump <- vector("list", 0) 
imp <- mice(data, m=5, meth="pmmdump", maxit=1, print=FALSE, seed=68006)
betadump <- matrix(betadump, nc=3, byrow=TRUE)
m1<-lm(Gas~Temp+Insul, data=data, na.action=na.omit)
an <- coef(m1)[1]
ai <- an + coef(m1)[3]
b <- coef(m1)[2]
abline(a=ai, b=b, col=mdc(4))
abline(a=an, b=b, col=mdc(4))
eta <- 0.6
ylo <- ai+b*(5-eta)
yhi <- ai+b*(5+eta)
lines(x=c(5-eta,5+eta),y=c(ylo,yhi),lwd=3,col=mdc(5))
xlo <- (ylo-an)/b
xhi <- (yhi-an)/b
lines(x=c(xlo,xhi),y=c(ylo,yhi),lwd=3,col=mdc(5))

donors <- subset(data, (Insul=="After"&Temp>5-eta&Temp<5+eta) 
                 |    (Insul=="Before"&Temp>xlo&Temp<xhi))
points(x=donors$Temp, y=donors$Gas, cex=1.8, col=mdc(5), lwd=lwd)
legend(x="bottomleft", legend=c("before insulation","after insulation"), pch=c(3,1),bty="n", pt.lwd=lwd)
text(x=9, y=6.5, label="f",cex=2)
```

### Section 3.2.3 Performance

```{r}
### number of simulations
### use nsim <- 10 for testing
###     nsim <- 10000 for the real stuff (TIME CONSUMING)
nsim <- 10

### create data
createdata <- function(beta0=5.49, beta1=-0.29, sigma=0.86, n=50, mx=5, sdx=3) { 
  x <- round(rnorm(n,mean=mx,sd=sdx),1)
  eps <- rnorm(n,mean=0,sd=sigma)
  y <- round(beta0 + x * beta1 + eps, 1)
  return(data.frame(x=x, y=y))
}

### make 50% random missing data
makemissing <- function(data, p=0.5){
  rx <- rbinom(nrow(data), 1, p)
  data[rx==0,"y"] <- NA
  return(data)
}

### test function for three imputation functions
test.impute <- function(data, m=5, method="norm", ...){
  imp <- mice(data, method=method, m=m, print=FALSE, ridge=0, ...)
  fit <- with(imp, lm(y~x))
  est <- pool(fit)
  tab <- summary(est)
  return(tab["x",c("est","se","lo 95","hi 95","fmi","lambda")])
}

### put everything together
simulate <- function(nsim=10, seed=41872){
  set.seed(seed)
  res <- array(NA,dim=c(4, nsim, 6))
  dimnames(res) <- list(c("predict","pred + noise","Bayes MI","boot MI"),
                      as.character(1:nsim),
                      c("est","se","lo 95","hi 95","fmi","lambda"))
  im <- c("norm.predict","norm.nob","norm","norm.boot")
  for(i in 1:nsim){
    data <- createdata()
    data <- makemissing(data)
    res[1,i,] <- test.impute(data, method=im[1], m=2)
    res[2,i,] <- test.impute(data, method=im[2])
    res[3,i,] <- test.impute(data, method=im[3])
    res[4,i,] <- test.impute(data, method=im[4])
  }
  return(res)
}

summarize.results <- function(res) {
  apply(res,c(1,3),mean,na.rm=TRUE)

  ## bias of beta1
  true <- -0.29
  bias <- rowMeans(res[,,1] - true)

  ## coverage
  isin <- res[,,3] < true & true < res[,,4]
  cov <- rowMeans(isin)

  ## average width of 95 c.i.
  intwidth <- res[,,4] - res[,,3]
  aiw <- rowMeans(intwidth)

  return(list(bias=bias, cov=cov, aiw=aiw))
}

### do the simulations - MAY BE TIME CONSUMING
res <- simulate(nsim)
summarize.results(res)
```

```{r}
### function to simulate performance of CCA
simulate.cca <- function(nsim=10, seed=41872){

  set.seed(seed)
  res <- array(NA,dim=c(1, nsim, 6))
  for(i in 1:nsim){
    data <- createdata()
    data <- makemissing(data)
    fit <- lm(y~x, data=data, na.action=na.omit)
    est <- coef(summary(fit))
    lo95 <- est[,1] - qt(0.975,fit$df) * est[,2]
    hi95 <- est[,1] + qt(0.975,fit$df) * est[,2]
    est <- cbind(est, lo95, hi95)
    res[1,i,] <- c(est["x",c(1:2,5:6)],NA,NA)
  }
  return(res)
}

### simulate - MAY BE TIME CONSUMING
res.cca <- simulate.cca(nsim)
### cannot be analysed by summarize.results() but it should
### be obvious how to calculate bias, coverage and aiw
```

### Section 3.2.4 Generating MAR missing data

```{r}
### script to create missing data according to MARRIGHT
logistic <- function(x) exp(x)/(1+exp(x))
set.seed(32881)
n <- 10000
y <- mvrnorm(n=n,mu=c(5,5),Sigma=matrix(c(1,0.6,0.6,1),nrow=2))
p2.marright <- 1 - logistic(-5 + y[,1])
r2.marright <- rbinom(n, 1, p2.marright)
yobs <- y
yobs[r2.marright==0, 2] <- NA

### Figure 3.2

grid <- seq(0,10,0.1)
mr <- logistic(-5 + grid)
mm <- logistic( 0.75-abs(grid-5))
mt <- logistic(-0.75+abs(grid-5))

par(mfrow=c(1,1))
z <- data.frame(grid, mr, mm, mt) 
matplot(x=z[,1],y=z[,2:4], type="l", lty=1:3, col=mdc(5) , lwd=2,
        xlab="Y1", ylab="Missingness in Y2", las=1)
legend(x="top", bty="n",
       legend=c("MARRIGHT","MARMID","MARTAIL"), lty=1:3, lwd=2, col=mdc(5), cex=0.8)
```

```{r}
### Figure 3.3

y3 <- rbind(y,y,y)
p2.marmid <- 1 - logistic(0.75-abs(y[,1]-5))
p2.martail <- 1 - logistic(0.75+abs(y[,1]-5))
r2.marmid <- rbinom(n, 1, p2.marright)
r2.martail <- rbinom(n, 1, p2.martail)
r2 <- c(r2.marright,r2.marmid,r2.martail)
r2 <- factor(r2, labels=c("Ymis","Yobs"))
typ <- factor(rep(3:1,each=n),labels=c("MARTAIL","MARMID","MARRIGHT"))
d <- data.frame(y1=y3[,1],y2=y3[,2],r2=r2,typ=typ)
trellis.par.set(box.rectangle=list(col=c(mdc(2),mdc(1)),lwd=1.2))
trellis.par.set(box.umbrella=list(col=c(mdc(2),mdc(1)),lwd=1.2))
trellis.par.set(plot.symbol=list(col="transparent",lwd=1))
tp <- bwplot(r2~y2|typ, data=d,
             horizontal=TRUE, layout=c(1,3),
             xlab=expression(Y2),
             col=c(mdc(2),mdc(1)),strip=FALSE, xlim=c(2,8),
             strip.left = strip.custom(bg="grey95"))
print(tp)
```

### Section 3.2.2 Imputation from the t-distribution

```{r}
mice.impute.tf <- function(y, ry, x, 
                  gamlss.trace = FALSE, ...)
{
  require(gamlss)

  # prepare data
  xobs <- x[ry, , drop = FALSE]
  xmis <- x[!ry, , drop = FALSE]
  yobs <- y[ry]
  n1 <- sum(ry)
  n0 <- sum(!ry)

  # draw bootstrap sample
  s <- sample(n1, n1, replace = TRUE)
  dotxobs <- xobs[s, , drop = FALSE]
  dotyobs <- yobs[s]
  dotxy <- data.frame(dotxobs, y = dotyobs)
  
  # fit the gamlss model
  fit <- gamlss(y ~ ., data = dotxy, family = TF, 
                trace = gamlss.trace, ...)
  yhat <- predict(fit, data=dotxy, newdata = xmis)
  sigma <- exp(coef(fit, "sigma"))
  nu <- exp(coef(fit, "nu"))
  
  # draw the imputations
  return(rTF(n0, yhat, sigma, nu))
}
```

### Section 3.3.3 Example

```{r}
data(db)

### Figure 3.4

par(mfrow=c(1,2))
data <- subset(db, age>1&age<2, c("age","head"))
names(data) <- c("age","hc")
truehist(data$hc, col=mdc(1), border="white", 
         xlab="Head circumference (cm)",
         ylab="Density",
         ylim=c(0,0.3), xlim=c(38,60),
         nbins=44)

mn <- gamlss(hc~1, data=na.omit(data), family=NO, trace=FALSE)
mu <- coef(mn)
sigma <- exp(coef(mn, "sigma"))
cmfine <- seq(38,60,0.1)
lines(cmfine, dNO(cmfine, mu, sigma), lwd=0.8, lty=2) 
mt <- gamlss(hc~1, data=data, family=TF, trace=FALSE)
mu <- coef(mt)
sigma <- exp(coef(mt, "sigma"))
nu <- exp(coef(mt, "nu"))
lines(cmfine, dTF(cmfine, mu, sigma, nu), lwd=0.8, lty=1) 
legend(x="right",legend=c("normal","t, df=6.7"), 
       lwd=0.8, lty=c(2,1), bty="n", cex=0.7)

plot(x=data$age, y=data$hc, col=mdc(1), cex=0.3,
     xlab="Age (in years)",
     ylab="Head circumference (cm)",
     ylim=c(39,60))
```

```{r}
### create a synthetic data set
data(db)
data <- subset(db, age>1&age<2, c("age","head"))
names(data) <- c("age","hc")
synthetic <- rep(c(FALSE,TRUE), each=nrow(data))
data2 <- rbind(data, data)
data2[synthetic,"hc"] <- NA
imp <- mice(data2, m=1, meth="tf", seed=36650, print=FALSE)
syn <- subset(complete(imp), synthetic)

### Figure 3.5

data <- syn
truehist(data$hc, col=mdc(2), border="white", 
         xlab="Head circumference (cm)",
         ylab="Density",
         ylim=c(0,0.3), xlim=c(38,60),
         nbins=44)

mn <- gamlss(hc~1, data=na.omit(data), family=NO, trace=FALSE)
mu <- coef(mn)
sigma <- exp(coef(mn, "sigma"))
cmfine <- seq(38,60,0.1)
lines(cmfine, dNO(cmfine, mu, sigma), lwd=0.8, lty=2) 
mt <- gamlss(hc~1, data=data, family=TF, trace=FALSE)
mu <- coef(mt)
sigma <- exp(coef(mt, "sigma"))
nu <- exp(coef(mt, "nu"))
lines(cmfine, dTF(cmfine, mu, sigma, nu), lwd=0.8, lty=1) 
legend(x="right",legend=c("normal",paste("t, df=",round(nu,1),sep="")), 
       lwd=0.8, lty=c(2,1), bty="n", cex=0.7)
plot(x=data$age, y=data$hc, col=mdc(2), cex=0.3,
     xlab="Age (in years)",
     ylab="Head circumference (cm)",
     ylim=c(39,60))
```

### Section 3.4.1 PMM, Overview

```{r}
### Figure 3.6

data <- boys[boys$age<=2,c("age","bmi")]
set.seed(87120)
data[sample(92:136,10),"bmi"] <- NA
imp <- mice(data, meth="norm", m=1, seed=32212, print=FALSE)
cd1 <- complete(imp)
imp <- mice(data, m=1, seed=32212, print=FALSE)
cd2 <- complete(imp)
r <- !is.na(data$bmi)
plot(data, col=mdc(1), xlab="Age", ylab="BMI")
points(cd1[!r,], col=mdc(2), pch=19, cex=0.8)
plot(data, col=mdc(1), xlab="Age", ylab="BMI")
points(cd2[!r,], col=mdc(2), pch=19, cex=0.8)
```

### Section 3.4.2 PMM, Computational details

```{r}
### Figure 3.7

par(mfrow=c(1,1))
data <- whiteside
lwd <- 1.5
data[47,"Gas"] <- NA

pch <- c(rep(3,26),rep(1,30))
plot(x=data$Temp, y=data$Gas, col=mdc(1), lwd=lwd, pch=pch, 
     xlab=expression(paste("Temperature (", degree, "C)")), 
     ylab="Gas consumption (cubic feet)")
betadump <- vector("list", 0) 
imp <- mice(data, m=5, meth="pmmdump", maxit=1, print=FALSE, seed=68006)
betadump <- matrix(unlist(betadump), nc=3, byrow=TRUE)
m1<-lm(Gas~Temp+Insul, data=data, na.action=na.omit)
an <- coef(m1)[1]
ai <- an + coef(m1)[3]
b <- coef(m1)[2]
abline(a=ai, b=b, col=mdc(4))
abline(a=an, b=b, col=mdc(4))
## for (i in 56:56) {
##    abline(a=unlist(betadump[i,1]), b=unlist(betadump[i,2]), col=mdc(5))
##    abline(a=unlist(betadump[i,1])+unlist(betadump[i,3]), b=unlist(betadump[i,2]), col=mdc(5))
## }
## points(rep(5,5),imp$imp$Gas, lwd=lwd, col=mdc(2), pch=20) 
eta <- 0.6
ylo <- ai+b*(5-eta)
yhi <- ai+b*(5+eta)
lines(x=c(5-eta,5+eta),y=c(ylo,yhi),lwd=3,col=mdc(4))
an <- 7.05; ai<-an-1.7; b <- -0.38
xlo1 <- (ylo-ai)/b
xhi1 <- (yhi-ai)/b
xlo2 <- (ylo-an)/b
xhi2 <- (yhi-an)/b
abline(a=an, b=b, col=mdc(5))
abline(a=ai, b=b, col=mdc(5))
lines(x=c(xlo1,xhi1),y=c(ylo,yhi),lwd=3,col=mdc(5))
lines(x=c(xlo2,xhi2),y=c(ylo,yhi),lwd=3,col=mdc(5))
abline(v=c(5-eta,5+eta),h=c(ylo,yhi),col=mdc(4),lty=3)
rect(xlo1,0,xhi1,8,col=hcl(0,100,40,0.05),border=NA)
rect(xlo2,0,xhi2,8,col=hcl(0,100,40,0.05),border=NA)
# abline(v=c(xlo1,xhi1,xlo2,xhi2),col=mdc(5),lty=3)

donors <- subset(data, (Insul=="After"&Temp>xlo1&Temp<xhi1) 
                 |    (Insul=="Before"&Temp>xlo2&Temp<xhi2))
points(x=donors$Temp, y=donors$Gas, cex=1.8, col=mdc(5), lwd=lwd)
legend(x="bottomleft", legend=c("before insulation","after insulation"), pch=c(3,1),bty="n", pt.lwd=lwd)
```

### Section 3.4.3 PMM, Algorithm 
```{r}
### simulate function used to calculate Table 3.3
simulate <- function(nsim=10, seed=41872){
  set.seed(seed)
  res <- array(NA,dim=c(3, nsim, 6))
  for(i in 1:nsim){
    data <- createdata()
    data <- makemissing(data)
    res[1,i,] <- test.impute(data, m=5)
    res[2,i,] <- test.impute(data, m=5, donors=1)
    res[3,i,] <- test.impute(data, m=5, donors=10)
  }
  return(res)
}

### and perform simulations - MAY BE TIME CONSUMING
res.pmm <- simulate(nsim)
summarize.results(res.pmm)
```

### Section 3.7.1

```{r}
### Figure 3.8

par(mfrow=c(1,2))
fit <- rpart(Gas ~ Temp + Insul, data=whiteside)
plot(fit, branch=0, margin=0.15)
text(fit, use=T,pretty=0,dig=3,cex=0.8)

leaf <- row.names(fit$frame)[fit$where]
label <- factor(leaf, labels=c(2,3,1,4,5))
plot(x=whiteside$Temp, y=whiteside$Gas, type="n",
     xlab=expression(paste("Temperature (", degree, "C)")), 
     ylab="Gas consumption (cubic feet)")
text(x=whiteside$Temp, y=whiteside$Gas, label=label, col=mdc(4), cex=0.6)
```


### Section 3.9.4 Converting selection and pattern-mixture models

```{r}
### Figure 3.9

par(mfrow=c(1,2))
s <- c(
       100, 0.015, 0.058, 0.02, 0.35,
       110, 0.024, 0.074, 0.03, 0.30,
       120, 0.043, 0.103, 0.05, 0.25,
       130, 0.091, 0.164, 0.10, 0.20,
       140, 0.145, 0.185, 0.15, 0.15,
       150, 0.307, 0.247, 0.30, 0.10,
       160, 0.157, 0.099, 0.15, 0.08,
       170, 0.107, 0.049, 0.10, 0.06,
       180, 0.055, 0.016, 0.05, 0.04,
       190, 0.033, 0.005, 0.03, 0.02,
       200, 0.023, 0.000, 0.02, 0.00,
       210, 0.023, 0.000, 0.02, 0.00
       )
sm <- matrix(s, nr=12, nc=5, byrow=TRUE)
snug <- 1.5
xx <- cbind(sm[,1]-5+snug,
        sm[,1]-5-snug,
        sm[,1]-5)
matplot(x=xx[,3], y=cbind(1-sm[,5]),
        col=mdc(1), type="p", lwd=2, lty=1, pch=20,
        xlab="Systolic BP (mmHg)",
        ylab="Observation probability")
matplot(x=xx, y=sm[,2:4], type="s",
        col=c(mdc(4),mdc(5),mdc(6)), lwd=2, lty=1,
        xlab="Systolic BP (mmHg)",
        ylab="Density")
```


## Chapter 4 Multivariate missing data

### Section 4.1 Missing data patterns

```{r}
library("mice")
library("lattice")

### define data sets pattern1-pattern4 with four patterns
data <- matrix(sample(1:100,4*8*3,replace=TRUE),nrow=8*4,
                dimnames=list(NULL,c("A","B","C")))
data <- as.data.frame(data)
data[c(31:32),"A"] <- NA
data[c(15:16,22:24,30:32),"B"] <- NA
data[c(6:8,12:16,17:21,27:29),"C"] <- NA
mdpat <- cbind(expand.grid(rec = 8:1, pat = 1:4, var = 1:3), r=as.numeric(as.vector(is.na(data))))
pattern1 <- data[1:8,]
pattern2 <- data[9:16,]
pattern3 <- data[17:24,]
pattern4 <- data[25:32,]

### Figure 4.1
types <-  c("Univariate","Monotone","File matching","General")
tp41 <- levelplot(r~var+rec|as.factor(pat), data=mdpat,
          as.table=TRUE, aspect="iso",
          shrink=c(0.9), 
          col.regions = mdc(1:2),
          colorkey=FALSE,
          scales=list(draw=FALSE),
          xlab="", ylab="",
          between = list(x=1,y=0),
          strip = strip.custom(bg = "grey95", style = 1,
            factor.levels = types))
print(tp41)
```

```{r fig.height=6}
md.pattern(pattern4)
p <- md.pairs(pattern4)
p


### proportion of usable cases
p$mr/(p$mr+p$mm)


### outbound statistics
p$rm/(p$rm+p$rr)

### Figure 4.2
par(mfrow=c(2,2))
fluxplot(pattern1, main="", xlim=c(-0.1,1.1), ylim=c(-0.1,1.1))
text(x=0.5,y=1,label="Univariate")
fluxplot(pattern2, main="", xlim=c(-0.1,1.1), ylim=c(-0.1,1.1))
text(x=0.5,y=1,label="Monotone")
fluxplot(pattern3, main="", xlim=c(-0.1,1.1), ylim=c(-0.1,1.1))
text(x=0.5,y=1,label="File matching")
fluxplot(pattern4, main="", xlim=c(-0.1,1.1), ylim=c(-0.1,1.1))
text(x=0.5,y=1,label="General")
```

```{r}
### calculate pobs, influx, outflux of general pattern
flux(pattern4)[,1:3]

### section 4.3 Monotone data imputation
### monotone data imputation on three columns
data <- nhanes2[,1:3]
md.pattern(data)
imp <- mice(data, visit="monotone", maxit=1, m=2)

### Numerical example, approximate two-step
ini <- mice(nhanes2, maxit=0)
pred <- ini$pred
pred["bmi","chl"] <- 0
pred["hyp",c("chl","bmi")] <- 0
imp <- mice(nhanes2, vis="monotone", pred=pred, maxit=1, m=2)
```

### Section 4.5 Fully Conditional Specification

```{r cache=TRUE}
### Example of slow convergence
generate <- function(n = c(1000, 4500, 4500, 0),
                     cor = matrix(c(1, 0.9, 0.9, 0.9, 1, 0.7, 0.9, 0.7, 1), nrow = 3)) {
  require(MASS)
  nt <- sum(n)
  cs <- cumsum(n)
  data <- mvrnorm(nt, mu = rep(0,3), Sigma = cor)
  dimnames(data) <- list(1:nt, c("X", "Y1", "Y2"))
  if (n[2] > 0) data[(cs[1]+1):cs[2],"Y1"] <- NA 
  if (n[3] > 0) data[(cs[2]+1):cs[3],"Y2"] <- NA
  if (n[4] > 0) data[(cs[3]+1):cs[4],c("Y1","Y2")] <- NA
  return(data)
}

impute <- function(data, m = 5, method="norm", print=FALSE, maxit = 10, ...) {
  statistic <- matrix(NA, nrow=maxit, ncol=m)
  for (iter in 1:maxit) {
    if (iter==1) imp <- mice(data, m = m, method = method, print=print, maxit = 1, ...)
    else imp <- mice.mids(imp, maxit=1, print=print, ...)
    statistic[iter,] <- unlist(with(imp, cor(Y1,Y2))$analyses)
  }
  return(list(imp=imp, statistic=statistic))
}

simulate <- function(ns=matrix(c(1000,500,250,100,50,0,
                               rep(c(4500,4750,4875,4950,4975,5000),2),
                       rep(0,6)), nrow=6), m = 5, maxit=10,
                       seed=1, ...) {
  if (!missing(seed)) set.seed(seed)
  s <- cbind(rep(1:nrow(ns), each=maxit*m),
             apply(ns, 2, rep, each=maxit*m),
             rep(1:maxit,each=m), 1:m, NA)
  colnames(s) <- c("k","n111","n101","n110","n100","iteration","m","rY1Y2")
  for (k in 1:nrow(ns)){
    data <- generate(ns[k,], ...)
    r <- impute(data, m=m, maxit=maxit, ...)
    s[s[,"k"]==k,"rY1Y2"] <- t(r$statistic) 
  }
  return(data.frame(s))
}

### perform simulation - TIME CONSUMING (10 minutes)
slow.demo <- simulate(maxit=150, seed=62771)
```

```{r cache=TRUE}
### Figure 4.3
labels <- c("90% missing", "95% missing", "97.5% missing", 
            "99% missing", "99.5% missing", "100% missing")
tp43 <- xyplot(rY1Y2~iteration|as.factor(k), group=m,
       data = slow.demo, layout=c(3,2),
       type="l", as.table = TRUE,
       ylab = "Correlation between Y1 and Y2",
       xlab = "Iteration", col=mdc(3),
       scales=list(y=list(alternating=1, tck=c(1,0))),
       strip = strip.custom(bg="grey95", style=1,
         factor.levels=labels))
print(tp43)
```

### Section 4.6.3 Illustration

```{r cache=TRUE}
## boys data: select subset
select <- with(boys, age>=8 & age<=21.0)

### version with all numeric data for jm and pmm
djm <- boys[select,-4]
djm$gen <- as.integer(djm$gen)
djm$phb <- as.integer(djm$phb)
djm$reg <- as.integer(djm$reg)

### version with categorical variables for fcs
dfcs <- boys[select,-4]

### impute according to joint multivariate normal
jm.10 <- mice(djm, method="norm", seed=93005, m=10, print=FALSE)

### impute according to predictive mean matching
pmm.10 <- mice(djm, seed=71332,m=10, print=FALSE)

### impute according to proportional odds model
fcs.10 <- mice(dfcs, seed=81420, m=10, print=FALSE)

### Figure 4.4
tp44 <- xyplot(jm.10, gen~age|.imp, subset=as.integer(.imp)<7, 
       xlab="Age", ylab="Genital stage")
print(tp44)
```

```{r}
### Figure 4.5
tp45 <- xyplot(fcs.10, gen~age|.imp, subset=as.integer(.imp)<7,
       xlab="Age", ylab="Genital stage")
print(tp45)
```

```{r}
### Figure 4.6 is too complex, and will not be given here
```

